Simulate Cut Off

### 2022-06-17 simulate_cutoff.Rmd
finalizing stimuli power

More real introduction on power, current trends to AIPE, and why items are key.

@@

Notes forthcoming.

@@

```{r echo = F}
# small potential variability overall, sort of 1-7ish scale
mu1 <- rnorm(30, 4, .25)
sigma1.1 <- rnorm(30, 2, .2)
sigma2.1 <- rnorm(30, 2, .4)
sigma3.1 <- rnorm(30, 2, .8)

@@

```{r}
# calculate the SEs and the cutoff scores 
SES1 <- SES2 <- SES3 <- list()
cutoffs1 <- cutoffs2 <- cutoffs3 <- list()
for (i in 1:length(samples1)){
  SES1[[i]] <- tapply(samples1[[i]]$score,
                     list(samples1[[i]]$item,
                          samples1[[i]]$scale),
                     function (x){ sd(x)/sqrt(length(x))})
  SES2[[i]] <- tapply(samples2[[i]]$score,
                   list(samples2[[i]]$item,
                        samples2[[i]]$scale),
                   function (x){ sd(x)/sqrt(length(x))})
  SES3[[i]] <- tapply(samples3[[i]]$score,
                 list(samples2[[i]]$item,
                      samples2[[i]]$scale),
                 function (x){ sd(x)/sqrt(length(x))})

@@

summary_long$source <- factor(summary_long$source, 
                              levels = c("low", "med", "high"),
                              labels = c("Low", "Medium", "High"))

summary_long$sd_original <- .2
summary_long$sd_original[summary_long$source == "Medium" &
                           summary_long$scale == "1"] <- .4
summary_long$sd_original[summary_long$source == "High" &
                           summary_long$source == "1"] <- .8

summary_long$sd_original[summary_long$source == "Low" &
                           summary_long$scale == "2"] <- 4
summary_long$sd_original[summary_long$source == "Medium" &
                           summary_long$scale == "2"] <- 8
summary_long$sd_original[summary_long$source == "High" &
                           summary_long$source == "2"] <- 12

summary_long$sd_original[summary_long$source == "Low" &
                           summary_long$scale == "3"] <- 100
summary_long$sd_original[summary_long$source == "Medium" &
                           summary_long$scale == "3"] <- 150
summary_long$sd_original[summary_long$source == "High" &
                           summary_long$source == "3"] <- 200

@@

```{r}
ggplot(summary_long %>% filter(power == 80), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ source) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

In our second question, we examined if the suggested procedure was sensitive to the amount of information present in the pilot data. Larger pilot data is more informative, and therefore, we should expect a lower projected sample size. As shown in the figure below for only the low variability data, we do not find this effect. These simplistic simulations from the pilot data would nearly always suggest a larger sample size - mostly in a linear trend increasing with sample sizes. This result comes from the nature of the procedure - if we base our estimates on some SE cutoff, we will almost always need a bit more people for items to meet those goals. This result does not achieve our second goal.

```{r}
ggplot(summary_long %>% filter(source == "Low"), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

$$ 1 - \sqrt{\frac{N_{pilot} - min(N_{sim})}{N_{pilot}}}^{log_2(N_{pilot})}$$
$N_{pilot}$ indicates the sample size of the pilot data minus the minimum sample size for simulation to ensure that the smallest sample sizes do not decay (e.g., the formula zeroes out). This value is raised to the power of $log_2$ of the sample size of the pilot data, which decreases the impact of the decay to smaller increments for increasing sample sizes. This value is then multiplied by the proposed sample size. As show in the figure below, this correction factor produces the dsired quality of maintaining that small pilot studies should *increase* sample size, and that sample size suggestions level off as pilot study data sample size increases.

@@

ggplot(summary_long %>% filter(source == "Low"), aes(original_n, new_sample, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

We have portrayed that this procedure, with a correction factor, can perform as desired. However, within real scenarios, researchers will only have one pilot sample, not many as show here. What should the researcher do to correct their sample size on their own pilot data?

To explore if we could recover the new suggested sample size from data a researcher would have, we used linear models to create a formula for calculation. First, the corrected sample size was predicted by the original suggested sample size. Next, the standard deviation of the item standard deviations was added to the equation. Last, we included the pilot sample size.

@@

user_model2 <- lm(new_sample ~ sample_size + sd_original, data = summary_long)
user_print2 <- apa_print(user_model2)

user_model3 <- lm(new_sample ~ sample_size + sd_original + original_n, data = summary_long)
user_print3 <- apa_print(user_model3)

user_model4 <- lm(new_sample ~ sample_size + sd_original + original_n + scale, data = summary_long)
user_print4 <- apa_print(user_model4)

@@

The first model using original sample size to predict new sample size was significant, `r user_print$full_result$modelfit$r2`, capturing nearly 90% of the variance. The second model with item standard deviation was better then the first model *F*(`r change_table$df[2]`, `r change_table$res.df[2]`) = `r change_table$statistic[2]`, *p* < .001, `r user_print2$estimate$modelfit$r2`. The addition of the original pilot sample size was also significant, *F*(`r change_table$df[3]`, `r change_table$res.df[3]`) = `r change_table$statistic[3]`, *p* < .001, `r user_print3$estimate$modelfit$r2`.

As shown in the table below, the new suggested sample size is proportional to the original suggested sample size (i.e., *b* < 1), which reduces the sample size suggestion. As variability increases, the suggested sample size also increases to capture differences in heterogeneity shown above. Last, in order to correct for large pilot data, the original pilot sample size decreases the new suggested sample size. This formula approximation captures 96% of the variance in sample size scores and should allow a researcher to estimate based on their own data.

@@

for (cutoff in unique(summary_long$name)){
  by_cutoff[[cutoff]] <- lm(new_sample ~ sample_size + sd_original + original_n, data = summary_long  %>% filter(name == cutoff))
  R2[cutoff] <- summary(by_cutoff[[cutoff]])$r.squared
}

@@

Last, we examine the question of an appropriate SE decile. All graphs for power, variability, and correction are presented below. If we examine the $R^2$ values for each decile of our regression equation separately, we find that the 10% (`r round(R2[["Decile 10"]], digits = 3)`) and 30% deciles (`r round(R2[["Decile 30"]], digits = 3)`) represent the best match to our corrected sample size suggestions. The 10% is likely unfeasible for many researchers, as the sample sizes are often quite large. The 30% decile, in the corrected format, appears to meet all goals: 1) increases with heterogeneity, 2) higher suggested values for lower samples and a leveling effect at larger pilot data, and ...

@@

```{r}
ggplot(summary_long %>% filter(source == "Low"), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

```{r}
ggplot(summary_long %>% filter(source == "Low"), aes(original_n, new_sample, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

```{r}
ggplot(summary_long %>% filter(source == "Medium"), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

ggplot(summary_long %>% filter(source == "Medium"), aes(original_n, new_sample, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

```{r}
ggplot(summary_long %>% filter(source == "High"), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

```{r}
ggplot(summary_long %>% filter(source == "High"), aes(original_n, new_sample, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

>>>>

@@

1) Take your original pilot data.
2) Find a critical score for an accurately measured item using SE.
3) Sample from your pilot data using sample sizes starting at 20 participants and increase in units of 5 up to a value you feel is the maximum sample size.
4) Calculate the number of items that meet your critical score at each sample size.
5) Find the smallest sample size for 80%, 85%, 90%, and 95% power. 
6) Designate your minimum sample size from these values, the stopping rule used from step 2, and the maximum sample size from these values.

@@

```{r}
# small potential variability overall, sort of 1-7ish scale
mu1 <- rnorm(30, 4, .25)
sigma1.1 <- rnorm(30, 2, .2)
sigma2.1 <- rnorm(30, 2, .4)
sigma3.1 <- rnorm(30, 2, .8)

@@

```{r}
# calculate the SEs and the cutoff scores 
SES1 <- SES2 <- SES3 <- list()
cutoffs1 <- cutoffs2 <- cutoffs3 <- list()
sd_items1 <- sd_items2 <- sd_items3 <- list()
for (i in 1:length(samples1)){
  sd_items1[[i]] <- samples1[[i]] %>% group_by(item, scale) %>% 
    summarize(sd = sd(score), .groups = "keep") %>% 
    ungroup() %>% group_by(scale) %>% summarize(sd_item = sd(sd))
  sd_items2[[i]] <- samples2[[i]] %>% group_by(item, scale) %>% 
    summarize(sd = sd(score), .groups = "keep") %>% 
    ungroup() %>% group_by(scale) %>% summarize(sd_item = sd(sd))
  sd_items3[[i]] <- samples3[[i]] %>% group_by(item, scale) %>% 
    summarize(sd = sd(score), .groups = "keep") %>% 
    ungroup() %>% group_by(scale) %>% summarize(sd_item = sd(sd))

  SES1[[i]] <- tapply(samples1[[i]]$score,
                     list(samples1[[i]]$item,
                          samples1[[i]]$scale),
                     function (x){ sd(x)/sqrt(length(x))})
  SES2[[i]] <- tapply(samples2[[i]]$score,
                   list(samples2[[i]]$item,
                        samples2[[i]]$scale),
                   function (x){ sd(x)/sqrt(length(x))})
  SES3[[i]] <- tapply(samples3[[i]]$score,
                 list(samples2[[i]]$item,
                      samples2[[i]]$scale),
                 function (x){ sd(x)/sqrt(length(x))})

@@

summary_long$source <- factor(summary_long$source, 
                              levels = c("low", "med", "high"),
                              labels = c("Low Variance", "Medium Variance", "High Variance"))

summary_long$scale2 <- factor(summary_long$scale, 
                              levels = c(1:3),
                              labels = c("Small Scale", "Medium Scale", "Large Scale"))

sd_items <- bind_rows(sd_items1, sd_items2, sd_items3)
sd_items$original_n <- rep(rep(sizes, each = 3), 3)
sd_items$source <- rep(c("Low Variance", "Medium Variance", "High Variance"), each = 3*length(sizes))
summary_long <- summary_long %>% 
  full_join(sd_items, 
            by = c("original_n" = "original_n", "scale" = "scale", 
                   "source" = "source"))

@@

```{r}
summary_long$source <- factor(summary_long$source, 
                              levels = c("Low Variance", "Medium Variance", "High Variance"))

ggplot(summary_long %>% filter(power == 80), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ source*scale2) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

In our second question, we examined if the suggested procedure was sensitive to the amount of information present in the pilot data. Larger pilot data is more informative, and therefore, we should expect a lower projected sample size. As shown in the figure below for only the low variability and small scale data, we do not find this effect. These simplistic simulations from the pilot data would nearly always suggest a larger sample size - mostly in a linear trend increasing with sample sizes. This result comes from the nature of the procedure - if we base our estimates on some SE cutoff, we will almost always need a bit more people for items to meet those goals. This result does not achieve our second goal.

```{r}
ggplot(summary_long %>% filter(source == "Low Variance") %>% filter(scale == 1), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

$$ 1 - \sqrt{\frac{N_{pilot} - min(N_{sim})}{N_{pilot}}}^{log_2(N_{pilot})}$$
$N_{pilot}$ indicates the sample size of the pilot data minus the minimum sample size for simulation to ensure that the smallest sample sizes do not decay (e.g., the formula zeroes out). This value is raised to the power of $log_2$ of the sample size of the pilot data, which decreases the impact of the decay to smaller increments for increasing sample sizes. This value is then multiplied by the proposed sample size. As show in the figure below, this correction factor produces the desired quality of maintaining that small pilot studies should *increase* sample size, and that sample size suggestions level off as pilot study data sample size increases.

@@

ggplot(summary_long %>% filter(source == "Low Variance") %>% filter(scale == 1), aes(original_n, new_sample, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

We have portrayed that this procedure, with a correction factor, can perform as desired. However, within real scenarios, researchers will only have one pilot sample, not the various simulated samples shown above. What should the researcher do to correct their sample size on their own pilot data?

To explore if we could recover the new suggested sample size from data a researcher would have, we used linear models to create a formula for calculation. First, the corrected sample size was predicted by the original suggested sample size. Next, the standard deviation of the item standard deviations was added to the equation to recreate heterogeneity estimates. Last, we included the pilot sample size. The scale of the data is embedded into the standard deviation of the items (*r* = `r round(cor(summary_long$scale, summary_long$sd_item), digits = 2)`), and therefore, this variable was not included separately. The standard deviation of the item's standard deviation embeds both heterogeneity and potential scale size.

@@

user_model2 <- lm(new_sample ~ sample_size + sd_item, data = summary_long)
user_print2 <- apa_print(user_model2)

user_model3 <- lm(new_sample ~ sample_size + sd_item + original_n, data = summary_long)
user_print3 <- apa_print(user_model3)

@@

The first model using original sample size to predict new sample size was significant, `r user_print$full_result$modelfit$r2`, capturing nearly 90% of the variance, `r user_print$estimate$sample_size`. The second model with item standard deviation was better then the first model *F*(`r change_table$df[2]`, `r change_table$res.df[2]`) = `r change_table$statistic[2]`, *p* < .001, `r user_print2$estimate$modelfit$r2`. The item standard deviation predictor was significant, `r user_print2$full_result$sd_item`. The addition of the original pilot sample size was also significant, *F*(`r change_table$df[3]`, `r change_table$res.df[3]`) = `r change_table$statistic[3]`, *p* < .001, `r user_print3$estimate$modelfit$r2`.

As shown in the table below for our final model, the new suggested sample size is proportional to the original suggested sample size (i.e., *b* < 1), which reduces the sample size suggestion. As variability increases, the suggested sample size also increases to capture differences in heterogeneity shown above; however, this predictor is not significant in the final model, and only contributes a small portion of overall variance. Last, in order to correct for large pilot data, the original pilot sample size decreases the new suggested sample size. This formula approximation captures 96% of the variance in sample size scores and should allow a researcher to estimate based on their own data.

@@

for (cutoff in unique(summary_long$name)){
  by_cutoff[[cutoff]] <- lm(new_sample ~ sample_size + sd_item + original_n, data = summary_long  %>% filter(name == cutoff))
  R2[cutoff] <- summary(by_cutoff[[cutoff]])$r.squared
}

@@

Last, we examine the question of an appropriate SE decile. All graphs for power, heterogeneity, scale, and correction are presented below. First, the 0%, 10%, and 20% deciles are likely too restrictive, providing very large estimates that do not always find a reasonable sample size in proportion to the pilot sample size, scale, and heterogeneity.If we examine the $R^2$ values for each decile of our regression equation separately, we find that the 50% (`r round(R2[["Decile 50"]], digits = 3)`) represents the best match to our corrected sample size suggestions. The 50% decile, in the corrected format, appears to meet all goals: 1) increases with heterogeneity and scale of data, and 2) higher suggested values for small original samples and a leveling effect at larger pilot data.

The formula for finding the corrected sample size using a 50% decile is: $Final Sample = 0.637 \times X_{Suggested Sample Size} + 0.002 \times X_{SD Items} + -0.546 \times X_{Pilot Sample Size}$.

## Conclusions

1) Take your original pilot data
2) Find a critical score for an accurately measured item using SE _*at the 50% decile of the item SE values*_.
3) Sample from your pilot data using sample sizes starting at 20 participants and increase in units of 5 up to a value you feel is the maximum sample size. 
4) Calculate the number of items that meet your critical score at each sample size.
5) Find the smallest sample size for 80%, 85%, 90%, and 95% power. 
6) _*Apply the correction to the smallest sample sizes*_.
7) Designate your minimum sample size from these values, the stopping rule used from step 2, and the maximum sample size from these values.

@@

```{r}
ggplot(summary_long %>% filter(source == "Low Variance"), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ scale2*power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

```{r}
ggplot(summary_long %>% filter(source == "Low Variance"), aes(original_n, new_sample, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ scale2*power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

```{r}
ggplot(summary_long %>% filter(source == "Medium Variance"), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ scale2*power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

ggplot(summary_long %>% filter(source == "Medium Variance"), aes(original_n, new_sample, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ scale2*power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

@@

```{r}
ggplot(summary_long %>% filter(source == "High Variance"), aes(original_n, sample_size, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ scale2*power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```

```{r}
ggplot(summary_long %>% filter(source == "High Variance"), aes(original_n, new_sample, color = name)) + 
  geom_point() +  
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_classic() + 
  facet_wrap(~ scale2*power) + 
  xlab("Original Sample Size") + 
  ylab("Suggested Sample Size") + 
  scale_color_discrete(name = "Cutoff Score")
```
